A Computational Account of Default-Mode Function\\by Control Theory and Reinforcement Learning

### 2016-07-16 paper.tex
fixed
added ToC
fixup
small stuff
new stuff

\section{Introduction to the Default-Mode Network}
%
scene construction processes when envisioning past, ficti- tious, and future events (Hassabis et al., 2007; Spreng et al., 2009).

@@

  \item
  autobiographical memory sup- plies numerous building blocks of social semantic knowl- edge (Bar 2007; Binder et al. 2009). These isolated conceptual scripts may be reassembled to enable the fore- casting of future events (Tulving 1983, 1985; Schacter et al. 2007). Similar brain mechanisms in remembering past episodes and envisioning future circumstances is supported by their engagement of identical brain areas, as evidenced by a quantitative meta-analysis (Spreng et al. 2009). Moreover, retrograde amnesic patients were repor- ted to be impaired not only in prospection but also in imagining novel experience (Hassabis et al. 2007). These findings suggest a single neural network for mentally constructing plausible semantic scenarios of detached sit- uations regardless of temporal orientation (Buckner and Carroll 2007; Hassabis and Maguire 2007). Indeed, con- struction of detached probabilistic scenes has been argued to influence ongoing decision making by estimating out- comes of behavioral choices (Boyer 2008; Suddendorf and Corballis 2007; Schilbach et al. 2008b). 
\end{itemize}

The Predictive Coding Framework.
\begin{itemize}
  \item It is one of the most frequently cited hypotheses of default mode function
  \item Within this model the brain is conceptualized as a Bayesian machine
  \item Predictive coding is a framework that, in a hierarchical setting,
  is equivalent to empirical Bayesian inference
  \item External sensory input is feed-forward/bottom-up processed in the brain
  \item Their processing is compared against expected input
  \item The prediction error, i.e. the difference between sensory input
  and internal prediction, is computed at each level and passed to
  higher levels via forward connections
  \item in case of a mismatch, back-projections top-down modulate
  input processing
  \item ‘contextual integration’ for top-down modulation of sensorimotor
  processing by context-specific a-priori information.
  \item Closely related to Friston's free-energy principle: brain as inference engine
  biological systems, including brains, must minimize the long-term average of surprise;
  formulate perception as a constructive process based on internal or generative models
  \item A generative model of the world
\end{itemize}

\section{Introduction to Control Theory and Reinforcement Learning}
\paragraph{Control Theory.}
\begin{itemize}
\item mathematically formalized dynamics systems with input, output and
feedback loops that produce error signals for adaptation
\item typically realized by (non-linear) differential equations
\item has been used to describe neural systems in the past
\item gradient update = learn/adapt system dynamics by error occurrence
\item related to notions of game theory
\item switching between states -> typically reduces to a sequential processing model
\item Tenenbaum2011Science: Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines (70–74).
\item policy: action to take given a state
\item value: prediction of future reward
\item model: predict what the environment will do next
\item model-baseed RI learning: we try to understand the environment; it is initially unknown
\item special to RL: exploitation vs. exploration
\item Tenenbaum2011Science: Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines (70–74).
\item Tenenbaum2011Science: Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines (70–74).
\item Tenenbaum2011Science: Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines (70–74).
\end{itemize}

@@

PFC
Patients with prefrontal lesions have problems in adapting to novel situations and stimuli
Stuss, D. T. & Benson, D. F. The Frontal Lobes (Raven, New York, 1986).

@@

\section{Relation to Other Cognitive Models (Danilo)}
\paragraph{Predictive Coding & Bayesian Brain Hypothesis}

\paragraph{Semantic Hypothesis}

\paragraph{Sentinel Hypothesis}

\section{Relation to Other Statistical Models (Elvis)}
\paragraph{Control Theory}

\paragraph{Q-Learning}

\paragraph{Free Energy Principle}

\section{Conclusion}

formalize and predict

>>>>

\tableofcontents

\section{The Default-Mode Network}
%
When left unperturbed, the human brain is not at rest. Rather, the brain continues to metabolize large quantities of oxygen and glucose energy to maintain inter-neuronal information transfer in the absence of an externally structured task (Bzdok and Eickhoff, 2016a). This baseline energy demand is subject to surprisingly little modulations due to the cognitive load posed by environmental challenges. What has early been described as the "stream of consciousness" in psychology (James, 1890).
%
A few years ago, the so-called "default mode network" was discovered entirely by accident (Shulman et al., 1997). A coherent set of brain regions consistently increased in neural activity during stimulus-independent thought, which was proposed to reflect the neural correlates underlying unfocused everyday mind wandering (Raichle et al., 2001). That is, in the beginning of the 21st century, brain imaging was the first technology to allow for the discovery of a coherent brain network that subserves a set of baseline mental activity (Bzdok et al., 2012; IBzdok et al., 2015).
%
15 years ago neuroscientists have discovered an energy-demanding brain network with unique properties. The "default-mode network" (DMN) appeared exclusive in task-induced deactivation during various psychological experiments. It was later argued to be systematically anti-correlated with brain regions that subserve task performance. Today, many authors speculate that the DMN implements continuous thinking about others’ and one’s own mind states, as well as adaptive envisioning of past, hypothetical, and future events. This network might have emerged to continuously predict environmental events using mental imagery as an evolutionary advantage.
%
Since 2001, the neurobiological properties of the default mode network have been investigated and reported in more than 3,000 neuroimaging publications 
The DMN was initially believed to represent the neural correlates of unconstrained mind-wandering because of its pervasively observed activity decrease during a large array of tasks. This processing of unknown information categories in the DMN has been argued to mediate an evolutionarily conserved function for the individual. This is all the more likely because the DMN contains the two biggest hotspots of energy consumption in the entire central nervous system. DMN activity also persists to a substantial degree during the early stages of sleep (Horovitz et al., 2008) and under anesthesia (Greicius et al., 2008). However, information processing in the DMN has also repeatedly been shown to impact human behavior. Goal-directed task performance improves with decreased activity in default-mode areas (Weissman et al., 2006) and increased DMN activity is linked to more task-independent, potentially disturbing thoughts (Mason et al., 2007). 
%
the DMN is likely to control the interplay between perception–action cycles and mental imagery (Bzdok et al., 2013a; Bzdok et al., 2013b). In particular, the DMN is likely to subserve a computational mechanism by its connections to highly associative brain areas that is key for human thought across time, splace, and content domains at the interface of external world and self (IBzdok et al., 2015).
%

scene construction processes when envisioning past, fictitious, and future events (Hassabis et al., 2007; Spreng et al., 2009).

@@

  \item
  autobiographical memory sup- plies numerous building blocks of social semantic knowl- edge (Bar 2007; Binder et al. 2009). These isolated conceptual scripts may be reassembled to enable the fore- casting of future events (Tulving 1983, 1985; Schacter et al. 2007). Similar brain mechanisms in remembering past episodes and envisioning future circumstances is supported by their engagement of identical brain areas, as evidenced by a quantitative meta-analysis (Spreng et al. 2009). Moreover, retrograde amnesic patients were repor- ted to be impaired not only in prospection but also in imagining novel experience (Hassabis et al. 2007). These findings suggest a single neural network for mentally constructing plausible semantic scenarios of detached sit- uations regardless of temporal orientation (Buckner and Carroll 2007; Hassabis and Maguire 2007). Indeed, con- struction of detached probabilistic scenes has been argued to influence ongoing decision making by estimating out- comes of behavioral choices (Boyer 2008; Suddendorf and Corballis 2007; Schilbach et al. 2008b). 
  \item
  DMN exists in monkeys [73] and rats [74]. 
  \item clinical research has corroborated default-mode network dysfunction in various psychiatric and neurological disorders
\end{itemize}

The DMN poses an unprecedented challenge to our neuroscientific understanding and methodological inventory.

Despite ramifications in health and disease, the functional role of the DMN poses an unprecedented challenge to the conventional interpretation strategies and existing methodological arsenal in the neurosciences.

Represents the physiological and pathological instantiation of a human beings' default mental repertoire, the nature of which remains largely obscure.

The DMN has now urged major attention in the neuroscientific community (Raichle et al., 2001, >6600 citations).

\section{Control Theory and Reinforcement Learning}
\paragraph{Control Theory.}
\begin{itemize}
\item mathematically formalized dynamics systems with input, output and
feedback loops that produce error signals for adaptation
\item typically realized by (non-linear) differential equations
\item has been used to describe neural systems in the past
\item gradient update = learn/adapt system dynamics by error occurrence
\item related to notions of game theory
\item switching between states -> typically reduces to a sequential processing model
\item Tenenbaum2011Science: Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines (70–74).
\item policy: action to take given a state
\item value: prediction of future reward
\item model: predict what the environment will do next
\item model-baseed RI learning: we try to understand the environment; it is initially unknown
\item special to RL: exploitation vs. exploration
\item Tenenbaum2011Science: Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines (70–74).
\item Tenenbaum2011Science: Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines (70–74).
\item Tenenbaum2011Science: Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines (70–74).
\end{itemize}

@@

PFC
Patients with prefrontal lesions have problems in adapting to novel situations and stimuli
Stuss, D. T. \& Benson, D. F. The Frontal Lobes (Raven, New York, 1986).

mechanistic relevance of each node -> we deconstruct

@@

\section{Relation to Other Cognitive Models (Danilo)}
\paragraph{Predictive Coding and Bayesian Brain Hypothesis}
abc
The Predictive Coding Framework.
construction of detached probabilistic scenes has been argued to influence ongoing decision-making by estimating outcomes of behavioral choices.

\begin{itemize}
  \item It is one of the most frequently cited hypotheses of default mode function
  \item Within this model the brain is conceptualized as a Bayesian machine
  \item Predictive coding is a framework that, in a hierarchical setting,
  is equivalent to empirical Bayesian inference
  \item External sensory input is feed-forward/bottom-up processed in the brain
  \item Their processing is compared against expected input
  \item The prediction error, i.e. the difference between sensory input
  and internal prediction, is computed at each level and passed to
  higher levels via forward connections
  \item in case of a mismatch, back-projections top-down modulate
  input processing
  \item ‘contextual integration’ for top-down modulation of sensorimotor
  processing by context-specific a-priori information.
  \item Closely related to Friston's free-energy principle: brain as inference engine
  biological systems, including brains, must minimize the long-term average of surprise;
  formulate perception as a constructive process based on internal or generative models
  \item A generative model of the world
\end{itemize}

\paragraph{Semantic Hypothesis}
abc

\paragraph{Sentinel Hypothesis}
many investigators speculate that this central neural network reflects the brain’s relentless tracking and prediction of environmental events to adaptively optimize the organism's future action (Schacter et al., 2007).

Cognitive neuroimaging studies linked the DMN to the contemplation of others’ (Mar, 2011) and one’s thoughts (Lombardo et al., 2009) mental states,

DMN does social cognition -> but it exists in rats and monkeys with much more poorly developped social capacities

How can a same neurobiological circuit be equally important for baseline house-keeping functions and specific task performance?

Our computational account can explain why the DMN is implicated in both a goal-directed task and an idlying rest cognitive set? -> task is policy/value updates to optmize short-term action / rest updates for mid- and long-term action

\section{Relation to Other Statistical Models (Elvis)}
\paragraph{Control Theory}
abc

\paragraph{Q-Learning}
abc

\paragraph{Free Energy Principle}
abc

\section{Conclusion}
What single brain function could be most important for existence and survival of the species?
From the perspective of the human condition, it remains unclear what the nature and purpose of a domain-overarching computational mechanism implemented in the high association cortex could be to warriant the high energetic costs.

; formalize and predict; offer statistical control on the brain's default function/operations
